---
title: "Tabula Muris: All FACS analysis"
output: html_notebook
---

# Preprocessing

Load the requisite packages and some additional helper functions.

```{r}
library(Seurat)
library(dplyr)
library(Matrix)
library(stringr)
library(readr)
library(here)


tabula.path ="/home/agranado/MEGA/Caltech/rnaseq/tabula-muris/"

FACS_files = list.files(paste(tabula.path,"00_data_ingest/", "00_facs_raw_data/FACS" ,sep=""), full.names = TRUE)

raw.data.list = list()
for (file in FACS_files){
  #read each file (csv count matrices sorted by tissue)
  # FACS_files[1] ="/home/agranado/MEGA/Caltech/rnaseq/tabula-muris/00_data_ingest/00_facs_raw_data/FACS/Aorta-counts.csv"
  raw.data <- read.csv(file, row.names = 1)
  raw.data <- Matrix(as.matrix(raw.data), sparse = TRUE)
  raw.data.list <- append(raw.data.list, raw.data)
}

#dim(raw.data.list[[2]])
#  [1] 23433  1638 # 1638 cells in this tissue, 23433 genes for ALL tissues]
# join the matrix with do.call cbind
raw.data <- do.call(cbind, raw.data.list)

  cell_order_FACS <- order(colnames(raw.data))
  raw.data = raw.data[,cell_order_FACS]

meta.data <- read.csv(paste(tabula.path,"00_data_ingest/", "00_facs_raw_data/metadata_FACS.csv",sep=""))
#we extract information from the colnames (ID of individual cells)

# meta.data includes the information for all plates

# > head(meta.data)
#           plate.barcode mouse.id  tissue subtissue FACS.selection mouse.sex
# D041914         D041914    3_8_M Bladder                 Multiple         M
# D042253         D042253    3_9_M Bladder                 Multiple         M
# MAA000487     MAA000487   3_10_M Bladder                 Multiple         M
# B000610         B000610   3_56_F Bladder                 Multiple         F
# B002764         B002764   3_38_F Bladder                 Multiple         F
# B002771         B002771   3_39_F Bladder                 Multiple         F

plates <- str_split(colnames(raw.data),"[.]", simplify = TRUE)[,2]

rownames(meta.data) <- meta.data$plate.barcode
#let's extract the plate information for all cells,
#plates is an array with the plate.id for each cell in the raw.data matrix
cell.meta.data <- meta.data[plates,]
rownames(cell.meta.data) <- colnames(raw.data) #lets rename the rows with the cell.ID
#because

# Find ERCC's, compute the percent ERCC, and drop them from the raw data.
# External RNA Controls Consortium
# The controls consist of a set of unlabeled, polyadenylated transcripts designed to be
# added to an RNA analysis experiment after sample isolation, in order to measure against defined performance criteria.

#look for more about ERCC here http://tools.thermofisher.com/content/sfs/manuals/cms_086340.pdf
erccs <- grep(pattern = "^ERCC-", x = rownames(x = raw.data), value = TRUE)
# there are genes (rows) that come from the ERCC control, so we can find them and calibrate the quantifitation
# there should be 92 ERCC transcripts

#percent is the ratio, for each cell, between the sum of ERCC detection divided by the total count
percent.ercc <- Matrix::colSums(raw.data[erccs, ])/Matrix::colSums(raw.data)
ercc.index <- grep(pattern = "^ERCC-", x = rownames(x = raw.data), value = FALSE)
raw.data <- raw.data[-ercc.index,] #remove the ERCC sequences

x11();hist(percent.ercc)
 # # # FILTER
raw.data = raw.data[,percent.ercc<0.15]
#filter on ERCC % which is a measure of RNA content. High ERCC means most of the sample comes from spike-in sequences

# # # # #HERE WE DO BMP-specific analysis

  # # # #Filter the bmp genes
  bmp.receptors<-c("Bmpr1a","Bmpr1b","Acvr1","Acvrl1","Acvr1b","Tgfbr1","Acvr1c","Acvr2a","Acvr2b","Bmpr2","Tgfbr2")
  bmp.ligands<-c("Bmp2","Bmp3","Bmp4","Bmp5","Bmp6","Bmp7","Bmp10","Bmp15",
              "Bmp8a","Gdf2","Gdf1","Gdf3","Gdf5","Gdf6","Gdf7","Gdf9","Gdf10","Gdf11","Gdf15")
  bmp.smads<-c("Smad1" ,"Smad2" ,"Smad3", "Smad4", "Smad5", "Smad6", "Smad7", "Smad9")

  bmp.all = c(bmp.receptors,bmp.ligands,bmp.smads)
  bmp.indexes=match(bmp.all,rownames(raw.data))

  raw.bmp <-raw.data[bmp.indexes,]

#take from meta.data only cells that remain in the matrix
cell.meta.filter = cell.meta.data[colnames(raw.data), ]
# Create the Seurat object with all the data


# # # # ## SEURAT OBJECT
tiss <- CreateSeuratObject(raw.data = raw.data)

tiss <- AddMetaData(object = tiss, cell.meta.data)
tiss <- AddMetaData(object = tiss, percent.ercc, col.name = "percent.ercc")
# Change default name for sums of counts from nUMI to nReads
colnames(tiss@meta.data)[colnames(tiss@meta.data) == 'nUMI'] <- 'nReads' # this is not UMI data so Seurat calculates only the number of reads

# > tiss
# An object of class Seurat
# 23341 features across 53760 samples within 1 assay
# Active assay: RNA (23341 features)

  ribo.genes <- grep(pattern = "^Rp[sl][[:digit:]]", x = rownames(x = tiss@data), value = TRUE)
  percent.ribo <- Matrix::colSums(tiss@raw.data[ribo.genes, ])/Matrix::colSums(tiss@raw.data)
  tiss <- AddMetaData(object = tiss, metadata = percent.ribo, col.name = "percent.ribo")

#re write for Seurat v3:
#ribo.genes <- grep(pattern = "^Rp[sl][[:digit:]]", x = rownames(tiss), value = TRUE)
#percent.ribo <- Matrix::colSums(tiss[ribo.genes, ])/Matrix::colSums(tiss)
#tiss <- AddMetaData(object = tiss, metadata = percent.ribo, col.name = "percent.ribo")


# #all of these percentages are for EACH cell
# percent.Rn45s <- tiss@raw.data['Rn45s', ]/Matrix::colSums(tiss@raw.data)
# tiss <- AddMetaData(object = tiss, metadata = percent.Rn45s, col.name = "percent.Rn45s")
```

# Analysis

A sanity check: genes per cell vs reads per cell.

```{r}
GenePlot(object = tiss, gene1 = "nReads", gene2 = "nGene", use.raw=T)

# # # # #Filter out cells with few reads and few genes.

#filter cells that have less that 50000 reads, or that have more than 500 genes with zero reads
tiss <- FilterCells(object = tiss, subset.names = c("nGene", "nReads"), low.thresholds = c(2, 50))
```

# # #Normalize the data, then center and scale.

```{r}
tiss <- NormalizeData(object = tiss, scale.factor = 1e6) #default normalization by Seurat
tiss <- ScaleData(object = tiss)
tiss <- FindVariableGenes(object = tiss, do.plot = TRUE, x.high.cutoff = Inf, y.cutoff = 0.5)
```

# # # #Run Principal Component Analysis.
```{r}
tiss <- RunPCA(object = tiss, do.print = FALSE, pcs.compute = 100)
tiss <- ProjectPCA(object = tiss, do.print = FALSE)
```

```{r, echo=FALSE, fig.height=4, fig.width=8}
PCHeatmap(object = tiss, pc.use = 1:3, cells.use = 500, do.balanced = TRUE, label.columns = FALSE, num.genes = 8)
```

Later on (in FindClusters and TSNE) you will pick a number of principal components to use. This has the effect of keeping the major directions of variation in the data and, ideally, supressing noise. There is no correct answer to the number to use, but a decent rule of thumb is to go until the plot plateaus.

```{r}
PCElbowPlot(object = tiss, num.pc = 5)
```

Choose the number of principal components to use.
```{r}
# Set number of principal components.
n.pcs = 5
```

The clustering is performed based on a nearest neighbors graph. Cells that have similar expression will be joined together. The Louvain algorithm looks for groups of cells with high modularity--more connections within the group than between groups. The resolution parameter determines the scale...higher resolution will give more clusters, lower resolution will give fewer.

For the top-level clustering, aim to under-cluster instead of over-cluster. It will be easy to subset groups and further analyze them below.

```{r}
# Set resolution
res.used <- 1

tiss <- FindClusters(object = tiss, reduction.type = "pca", dims.use = 1:n.pcs,
    resolution = res.used, print.output = 0, save.SNN = TRUE) #DONE
```

To visualize
```{r}
# If cells are too spread out, you can raise the perplexity. If you have few cells, try a lower perplexity (but never less than 10).
tiss <- RunTSNE(object = tiss, dims.use = 1:n.pcs, seed.use = 10, perplexity=30)
```

```{r}
TSNEPlot(tiss, group.by = 'tissue')
```

```{r}
save(tiss, file=here("00_data_ingest", "11_global_robj", "FACS_all_preannotation.Robj"))
#load(here("00_data_ingest", "11_global_robj", "FACS_all_preannotation.Robj"))
```

# Add in metadata, annotations, and save.

```{r}

# plot the location of a list of genes in the main tSNE plot
x11();FeaturePlot(tiss,bmp.ligands, cols.use = c("lightgrey","blue"),nCol=4  )

tiss@meta.data['cluster'] <- tiss@ident
tiss@meta.data['cell'] <- rownames(tiss@meta.data)
```

```{r}
#anno = read_csv(here("00_data_ingest", "18_global_annotation_csv", "annotations_FACS.csv"))
anno = read_csv(here("00_data_ingest", "18_global_annotation_csv", "annotations_facs.csv"))
anno %>% group_by(tissue, cell_ontology_class) %>% summarize(count = n())
```

```{r}
tissue_colors = read_csv(here("00_data_ingest", "15_color_palette","tissue_colors.csv"))
colnames(tissue_colors) = c("tissue", "color")
```


```{r}
tiss@meta.data <- tiss@meta.data %>%
		   left_join(anno %>% select(cell_ontology_class,cell_ontology_id,free_annotation, cell), by = 'cell') %>%
		   left_join(tissue_colors, by = 'tissue')

rownames(tiss@meta.data) <- tiss@meta.data$cell
```

```{r}
tiss_FACS = tiss
save(tiss_FACS, file=here("00_data_ingest", "11_global_robj", "FACS_all.Robj"))
```
